Getting off the ground: bootstrap learning via modular concept discovery.

### 2013-01-25 drafts/ijcai2013/ijcai13.tex
Changes.

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage{tikz}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata, positioning}

@@

\title{The E-C Algorithm: conceptual bootstrapping via grammar-based
  program compression.}

\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Josh Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
\And 
Ryan Adams \\
Harvard University\\
USA \\
rpa@seas.harvard.edu}

@@

\begin{abstract}
Suppose a learner is faced with a set of problems about which it knows
nearly nothing: it doesn't know the distribution of problems, or what
an appropriate loss function might be. For each hypothesized solution
to a problem it gets a few bits of positive or negative reinforcement,
but not enough to steer it effectively in the huge space of possible
solutions. How can such a learner every get off the ground? The key is
to find, by brute force, a solution to one or more simple problems in
that domain and use the information thus gained to pursue subsequent
searches in an informed manner. Here, we formalize this intuition,
where the solution space is the space of typed functional programs and
the gained information is stored as a stochastic grammer over
programs. Specifically, we propose the E-C algorithm, a two step
iterative procedure for exploring such spaces: in the E step, the
learner \emph{explores} a finite subset of the domain for solutions,
guided by a stochastic grammar; in the C step, the learner
\emph{compresses} the successful solutions from the E step to estimate
a new stochastic grammar. As we show in several toy domains, iterating
these two steps quickly propels the learner from a state of solving
almost no problems to solving a large number of problems by gaining
abstract knowledge of the structure of the solution space.
\end{abstract}

\section{Introduction.}

Suppose that, having been shipwrecked and saved by a people with no
knowledge of your language or culture, you find yourself attempting to
communicate in a language about which you know nothing. How would you
go about discovering even the most basic mappings of sounds to
meanings? Overhearing the native speech, you have a sense of what the
phonetics might be like, so you might try various common combinations
of sounds while pointing to things you want. Mostly, you receive
puzzled looks, but if you are lucky, you might pick up on a few key
words or phrases, like ``that'' or ``I want.'' Naturally, you will
proceed to use your discoveries to accelerate your learning, mixing
and matching known phrases to generate new meanings.

\section{Related work.}

\section{A multi-task program induction problem.}
Our goal is to solve a general multi-task learning problem: we are
given set of \emph{tasks} $T=\{t_k\}_{k=1}^K$ where each task is a
function $t_i : \mathcal{L} \rightarrow \mathds{R} > 0$ which returns
a reward value for every element in the solution representation
language $\mathcal{L}$. Our goal is to find a set of solutions
$X = \{x_k\}_{k=1}^k \subset \mathcal{L}$ such that

@@

\section{A compression scheme expresses a belief about the data.}
There is a deep connection between a distribution over data and the
optimally compressive code for that data. In fact, if we know the
probability distribution for a set of observations, we can obtain an
optimally compressive encoding of those observations via arithmetic
coding. Conversely, the encoding generated by a good compression
scheme can be interpreted as an approximation of the data
distribution.

Take, for example, the Nevill-Manning algorithm for grammar based
compression~\cite{nevill1997identifying}. This algorithm looks for
hierarchical structure in the data and represents that hierarchical
structure as a grammar. The output of the algorithm can be seen as a
grammar and a corresponding parse of the input data. Given that parse,
we can estimate the maximum likelihood probabilistic context-free
grammar for that data~\cite{johnson1998pcfg}.

\section{The E-C Algorithm.}

In this section, we describe the E-C algorithm and discuss its uses,
applicability, and limitations.

As its name suggests, each iteration of the E-C algorithm consists of
an exploration step and a compression step. In the exploration step,
we enumerate the $N$ most probable hypotheses given our currenst
distribution $D$ over hypotheses. We then assign to each task $t_i$
the hypotheses that match it. We say that a task is \emph{hit} if
there is a least one hypothesis that matches it. In the compression
step, our goal is to assign a single hypothesis $h_i^*$ to each task
$t_i$ such that that resulting set of hypotheses $\{t_i\}$ is as
compressible as possible. The compression this process induces is used
to update distribution $D$.

  \begin{algorithm}
    \SetAlgoLined \KwData{Tasks $T=\{t_1, \dots, t_K\}$,
      distribution $\mathcal{D}_j$ over expressions.}

    $\text{types} \leftarrow \cup_k \, type(t_k)$. \\
    For $\tau \in \text{ types}$ , $H_N(\tau) \leftarrow$ enumerate
    $N$ most probable elements in $\mathcal{D}_t$ with type $\tau$.\\

    For each \emph{hit} $s_i$ in 
    $T$, $v_{s_i} \leftarrow \{h \in D_j | s_i(h) == True \}$.\\

    $S = \{(s_i, h^*_{s_i})\} \leftarrow$ assignment of elements to tasks s.t. 
    $S$ is most compressible. \\

    $D_{j+1} \leftarrow$ derive new distribution from compression of $S$. \\

    Repeat.

    \caption{E-C}
  \end{algorithm}

\section{Our implementation.}

\subsection{Representing polymorphically typed programs as binary trees.}

Following~\cite{liang10programs} and~\cite{Briggs:2008}, we use
combinatory logic -- a variable-free subset of the polymorphic
simply-typed lambda calculus lambda calculus~\cite{Pierce_2002} -- as
our program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function in the lambda calculus can be alternatively written
as applications of the basis combinators. Some common basis
combinators are defined as follows:
\begin{align}
I\; x &\rightarrow  x \\
S\; f\; g\; x\; &\rightarrow (f\; x)\; (g\; x) \\ 
B\; f\; g\; x\; &\rightarrow (f\; x)\; g \\ 
C\; f\; g\; x\; &\rightarrow f\; (g\; x) \\ 
\end{align}

The basis combinators can
themselves been expressed in the lambda calculus. The lambda calculus
has two basic operation -- application and abstraction -- but in using
the combinatory logic we sequester uses of the abstraction operation
inside the combinators, and thus have a representation language that
is effectively abstraction- and, thus, variable-free.

@@

As a simple example of our representation, consider the squaring
function $\lambda x. * x\; x $. Using two of the basis combinators above,
we can write the squaring function in combinatory logic as $S * I$,
where application associates to the left, so we drop the
parentheses. When we apply this combinator to a value $x$, the action
of the combinator is defined as $S \;*\; I\; x~\rightarrow~(*\; x)\;
(I\; x)~\rightarrow(*\; x)\; x$. Now, we can extend this
representation with a simple polymorphic type
system~\cite{Pierce_2002}. Thus, this combinator can be represented as
a binary tree~(Figure~\ref{fig:clbintree}) whose leaves are typed
primitive combinators and whose interior nodes are typed applications
of combinators.

\begin{figure}
  \centering
  \begin{subfigure}[Before]{0.5\linewidth}
    \begin{tikzpicture}[scale=.9,
        transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$I$}
      child {node [above=.3cm] (a) {$(I \rightarrow I) \rightarrow I \rightarrow I$}
        child {node [left=.5cm, below=1cm, text width=3cm] 
          {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
      }
      child {node [below right=.3cm and 1cm] (b) {$I \rightarrow I$}
        child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };
    \end{tikzpicture}
    \label{fig:clbintree}
  \end{subfigure}
  \begin{subfigure}[After]{0.4\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}
    \label{fig:andor}
    \end{subfigure}
  \caption{(a) The typed combinator $S * I$ ($f(x)=x^2$) represented as
    binary tree. (b) The space of all typed combinators represented as an
    infinitely deep AND/OR tree. Any specific combinator as a
    sub-binary tree of this And-Or tree whose leaves consist of only
    the $f_i$ nodes (i.e., the primitive combinators of the grammar).}

@@

\subsection{Best-first enumeration of most probably programs.}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. We can
formulate this naturally as a best-first exploration of an $AND/OR$
tree~\cite{nilsson1982principles}, which we denote $\mathcal{G}$. A
complete program is any binary subtrees of $\mathcal{G}$ whose leaves
coincide with the leaves of $\mathcal{G}$. The tree structure is shown
in Figure~\ref{fig:andor}. The root is an $OR$ node of the requisite
program type. It's children are the primitives of that type $f_1,
\dots, f_n$, and one $AND$ node with two children. Each child of an
$AND$ node has the same structure as the root node, with modified
types: if the type of an $AND$ node is $t$, then the type of its left
child is $\tau \rightarrow t$ (where $\tau$ is a fresh type
variable). Since the type of the right child is constrained by the
subtree rooted at the left child, we always expand the left child
first and use the target type of this left subtree to mark the type of
the right child.

In accordance with the stochastic grammar over programs, we weight all
edges from $OR$ nodes according to the negative log likelihood of
choosing the associated rule from the stochastic grammar, conditioned
on the requesting type. Expanding an $AND$ node has zero cost. In the
case that the stochastic grammar is ill condition, we might have an
infinitely deep solution tree whose cost does not increase as we
unfold it; this can occur, for example, if there are no primitives of
the required type since the log probability of expansion to the next
$AND$ node is 0. . To avoid this, we simply declare a cost for
expanding the next $AND$ node when it is the only node available.

\subsection{Finding the most compressive set of solution.}

@@

In our algorithm, we traverse a set of combinator binary trees in
depth first order. We count the occurences of every subtree. Every
time we encounter a subtree for the \emph{second} time, we designate
it as a primitive in our new grammar. This ensures that every
primitive is used more than once. During our traversal, we do not
descend into trees that have been added already to the new
grammar. Thus, we only count an instance of tree if that instance is
not a subtree of a primitive element of the new grammar.

At the end of this traversal of the solution programs, we have a parse
of the solution set with a new set of primitives and we know how many
times each primitive occured in the solutions et. To estimate the
probabilities associated with each node production, we again traverse
the solution set, but this time for each node we count which other
primitive elements from the new grammar could have been used. We
estimate the terminal production weight to be the ratio of the number
of times that node was used in the first traversal divided by the
number of times it was used in the second traversal.

@@

The choice of an initial grammar for the algorithm will, of course,
have a big effect on its performance, and so we would like to
characterise this effect. We can divide primitives into roughly two
sets; there are low-level primitives, those that operate on concrete
types like Bool or Double, and these tend to be domain specific. For
example, conrete primitives include $1, 2$, etc., and $f(x) = 3x$. The
other class of primitives are abstract primitives, which can be
characterized by having arguments whose types are primarily function
types, like $(t \rightarrow s), (t \rightarrow s) \rightarrow \text{
  Double },$ etc. Of course, we can construct operators that fall in
between.

The most abstract operators are those that Liang et. al refer to as
``routers''~\cite{liang10programs}. These operators take the place of
$\lambda$ abstraction in the general lambda calculus: whereas
$\lambda$ abstraction allows us to bind arguments to variables nested
deep inside an expression, without them we need to have operators that
``route'' arguments to the appropriate functions in a nested
expression.

\begin{definition}[Router~\cite{liang10programs}] A \emph{router} is a combinator
represented by a finite sequence of elements from $\{B, C, S\}$. The
behavior of router $r$ is given by
\[
(r \, f \, g \, x_1 \dots x_{|r|}) = (f \, x_{i_1} \dots x_{i_n}) (g
\, x_{j_1} \dots x_{j_n}),
\]
where $i_1 < \dots < i_n$ are indices $i$ such that $r_i \in \{C, S\}$
and where $j_1 < \dots < j_n$ are indices $j$ such that $r_j \in \{B,
S\}$. Let $\mathcal{R}_k$ be the set of all routers of length equal to
$k$. Let $\mathcal{R}_{\leq k}$ be the set of all routers of length
less than or equal to $k$.
\end{definition}

To analyze the behavior of our algorithm, we will vary the parameters
of the problem along two axes while keeping the learning tasks the
same. Along the first axis, we vary the initial library to which our
learner has access with respect to the its abstractness (i.e., number
of routers) and its concreteness (i.e. number of concrete
primitives). Along the second axis we vary the size of the search
frontier.

Each task is to predict the value of an unknown polynomial on the
inputs $0, \dots, 10$. The set of tasks corresponds to the set of
polynomials $\{ax^2 + bx + c | a, b, c \in \{0, \dots, 10 \}\}$.

In Figure~\ref{fig:symreg}, we show performance results as we vary the
frontier size. Each individual plot shows how changing the frontier
size changes the number of tasks the algorithm identifies correctly
over the course of fifteen algorithm iterations. In addition, we have a
baseline for each curve, visits the same number of expressions as the
corresponding algorithm run (i.e. frontier size $\times$ $\#$ of
iterations) but all at once; it does not do any learning. Thus, this
baseline serves as a brute-force control for the learning aspect of
the algorithm.

\begin{figure}
\begin{subfigure}[Before]{0.45\linewidth}
\includegraphics[width=\linewidth]{figures/learningCurves2.pdf}
\end{subfigure}
\begin{subfigure}[After]{0.45\linewidth}
\includegraphics[width=\linewidth]{figures/performanceVsFrontierSize2.pdf}
\end{subfigure}
\caption{E-C learning curves for the symbolic regression task set as
  we vary the frontier size.}
\label{fig:symreg} 
\end{figure}

Why do the learning curves asymptote early? One might object that a
learning algorithm should continuously improve, if only slowly, as it
is given more time. But in these runs we observe a finite and
permanent performance cap. This is expected: once the learner find a
set of concepts it can explain, if those concepts fill up the
frontier, then there is nothing to spur the learner to incorporate as
yet unsolved tasks.

To investigate whether this is in fact the case, we can examine the
proportion of expressions in the frontier that match tasks in our task
set; that is, is the learning curve plateauing because the frontier is
heavily redundant, or is it plateuing because the frontier contains
many expressions that do not hit the task set at
all. Figure~\ref{fig:plateau_investigation} shows the proportion of
the expressions in the frontier that hit a task as the frontier sizeincreases.

>>>>

\usepackage{ijcai13}
\usepackage{times}
\usepackage{amsmath} 
\usepackage{latexsym} 
\usepackage{ dsfont }
\usepackage{tikz}
\usepackage{circuitikz}
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{alltt}
\usepackage[ruled,vlined,linesnumbered]{algorithm2e}
\usetikzlibrary{arrows,automata, positioning, patterns}

@@

\title{Getting off the ground: bootstrap learning via modular concept discovery.}

\author{Eyal Dechter \\
MIT\\
USA \\
edechter@mit.edu
\And
Jon Malmaud \\
MIT \\ 
USA \\
malmaud@mit.edu
\And 
Ryan P.  Adams \\
Harvard University\\
USA \\
rpa@seas.harvard.edu
\And
Josh Tenenbaum \\
MIT\\
USA \\
jbt@mit.edu
}

@@

\begin{abstract}
 Suppose a learner is faced with a domain of problems about which she
 knows nearly nothing. She does not know the distribution of problems,
 the space of solutions is not smooth, and the reward signal is
 uninformative, providing perhaps a few bits of information but not
 enough to steer her effectively. How can such a learner ever get off
 the ground? A common intuition is that if the solutions to these
 problems share a common structure, and the learner can solve some
 simple problems by luck or brute force, she should be able to extract
 useful modular concepts from these solutions and, by composing them,
 explore the solution space more efficiently. Here, we formalize this
 intuition, where the solution space is the space of typed functional
 programs and the gained information is stored as a stochastic grammar
 over programs. We propose a two step iterative procedure for
 exploring such spaces: in the first step, the learner explores a
 finite subset of the domain, guided by a stochastic grammar; in the
 second step, the learner compresses the successful solutions from the
 first step to estimate a new stochastic grammar. As we show in a
 symbolic regression and a boolean circuit learning example, iterating
 these two steps quickly propels the learner from a state of solving
 almost no problems to solving a large number of problems by gaining
 abstract knowledge of the structure of the solution space.
\end{abstract}

\section{Introduction}

Imagine you are trying to learn to bake via trial and error. You know
what kind of ingredients are used in baking (flour, water, butter,
sugar, yeast, baking soda, etc.) and you know a bit about how to
combine these ingredient (mix, whisk, etc.) and how to apply heat
(oven). But baking is a notoriously fragile process: ratios of
ingredients, the order of mixing, cooking temperatures, resting times,
all make a huge difference, and the outcome of the baking process is
mostly all or none. Either the outcome is what you expected or quite
unlike it. Even if you do create something similar to your intent, it
is difficult to assign blame for the outcome to any one step or
decision along the way.

Many difficult learning problems have this structure -- learning
language, developing technologies and scientific theories, learning to
play instruments or execute other complex motor patterns. These are
problems in which 1) there is a large set of similar tasks; 2) their
solutions are best described as programs; 3) the solution space is
not smooth with respect to the outcome; 4) and the outcome is generally
uninformative unless you are already pretty close to a solution.

Intuitively, the way to solve such problems is to stumble on solutions
to simple cases in the domain and extract conceptual building blocks
that can be composed to scale up to complex cases. To return to the
baking example, cookies are much simpler than cakes. There are fewer
steps to get just right. But once you have figured out a few simple
pastries, the more complex ones are mostly just variations and
combinations of techniques learned along the way.

Of course, knowledge is not acquired wholly by trial and error. But in
many domains, there is at least initially such a period of ignorance,
and the question we are trying to tackle here is how you ever get off
the ground.

To frame this learning paradigm computationally, we make two modeling
choices. First, to emphasize the impoverished reward signals in these
domains, we model the learning task as an all-or-none optimization
problem: either the learner gets some reinforcement or she does
not. Second, to emphasize the lack of locally smooth structure in the
space of possible solutions, we choose as our solution space the space
of functional programs. These two choices are in contrast to the
paradigm tackled by most of machine learning, in which a locally
informative objective function guides search in a space that is
somewhat smooth.

Formally then, our goal is to solve the following minimal feedback
multi-task program induction problem: suppose we are
given set of \emph{tasks} $T=\{t_k\}_{k=1}^K$ where each task is a
function $t_i : \mathcal{L} \rightarrow \{0, 1\}$, where $\mathcal{L}$
is the set of programs in our language.  We say that $ p \in
\mathcal{L}$ \emph{solves} $t_i$ if $ t_i\; p = 1$. Our
goal is to find a set of solutions $X = \{x_k\}_{k=1}^k \subset
\mathcal{L}$ such that

@@

Based on the intuition set out above, we present an iterative
strategy. Given a stochastic grammar $\mathcal{G}$ which defines a
distribution over the primitives from $\mathcal{L}$ and a integer
\emph{frontier size} $N$, we perform to steps in each iteration.  In
the first step, we explore the $N$ highest probability programs
according $\mathcal{G}$. In the second step, we compress the
programs that solve any of our tasks, and we use that compression to
define a new grammar. Repeating these two steps, we effectively
reparameterize our search space to include programs that are more likely
to be solutions to our tasks.

The remainder of this paper will include three major sections. First,
we will describe the representation language we chose and the goals of
the exploration and compressions steps. Following that, we will
discuss the algorithmic details that render our procedure
feasible. Finally, we present results from two experiments, involving
the domains of symbolic regression and boolean circuit learning.

%% \section{}
%% Let $\mathcal{F}$ be a family of distributions over $\mathcal{L}$ such
%% that for any $\mathcal{D} \in \mathcal{F}$ we can a) evaluate the
%% probability of program $p$, $P_D(p)$, and we can b) enumerate elements
%% from $D$ in descending order.

%%  To motivate a solution to our problem, we begin with the hierarchical
%% Bayesian formulation of multi-task program induction found in Liang et
%% al.~\cite{liang10programs}. The general idea there is to estimate
%% solution programs for a set of related problems by inferring the posterior
%% distribution over solutions under the assumption that all the solution
%% programs are generated from a common latent stochastic grammar. By
%% jointly inferring solutions and the latent grammar, we find a set of
%% solutions that favors reuse of subcomponents.

%% While this framework is appealing in its elegance and Bayesian
%% interpretation, it cannot be straightforwardly adapted to the
%% situation we are considering here because their proposed MCMC
%% algorithm relies on a local log-likelihood signal, the very kind that
%% we are assuming we do not have access to.

%% To begin, we provide a high level overview of the algorithm and the
%% problem that it solves. The problem setup requires a representation
%% language for programs and a family of distributions over expressions
%% in that language, but we leave these details for the following section

\section{Extracting modular concepts from functional programs}

In order to solve our stated problem, we need to choose a solution
representation language and a compression criterion, and we need to
describe the frontier enumeration and compression procedures.

\subsection{Representing typed functional programs as binary trees.}

Following~\cite{liang10programs} and~\cite{Briggs:2008}, we use
combinatory logic -- a variable-free subset of the polymorphic
simply-typed lambda calculus~\cite{Pierce_2002} -- as
our program representation language. In short, the combinatory logic
introduces a \emph{basis} of several primitive \emph{combinators} such
that any function in the lambda calculus can be alternatively written
as applications of the basis combinators. Some common basis
combinators are defined as follows:
\begin{align}
I\; x &\rightarrow  x \\
S\; f\; g\; x\; &\rightarrow (f\; x)\; (g\; x) \\ 
B\; f\; g\; x\; &\rightarrow (f\; x)\; g \\ 
C\; f\; g\; x\; &\rightarrow f\; (g\; x) \\ 
\end{align}

The basis
combinators are theoretically enough to express any Turing computable
function, but we will assume that our language has a variety of
primitives and, together with the basis combinators, we call these the
primitive combinators.

The basis combinators can themselves been expressed in the lambda
calculus. The lambda calculus has two basic operation -- application
and abstraction -- but in using the combinatory logic we sequester
uses of the abstraction operation inside the combinators, and thus
have a representation language that is effectively abstraction- and,
thus, variable-free. In doing so, we have replaced the variable
binding of the $\lambda$ operator with the variable \emph{routing} of
these basis combinators. See~\cite{liang10programs} for a more
detailed discussion of this routing interpretation.

@@

As a simple example of our representation, consider the squaring
function $\lambda x. * x\; x $. Using two of the basis combinators
above, we can write the squaring function in combinatory logic as $S *
I$ (taking $*$ as a primitive), where application associates to the
left, so we drop the parentheses. When we apply this combinator to a
value $x$, the action of the combinator is defined as $S \;*\; I\;
x~\rightarrow~(*\; x)\; (I\; x)~\rightarrow(*\; x)\; x$.

We can extend this representation with a simple polymorphic type
system~\cite{Pierce_2002}. Thus, this combinator can be represented as
a binary tree~(Figure~\ref{fig:clbintree}) whose leaves are typed
primitive combinators and whose interior nodes represent typed applications
of combinators.

\begin{figure}
  \centering
  \begin{subfigure}[Before]{0.5\linewidth}
    \begin{tikzpicture}[scale=.9,
        transform shape, level/.style={sibling distance=10mm/#1, font=\scriptsize}
        ]
      \node [font=\scriptsize] (top) {$Int$}
      child {node [above=.3cm, text width=2cm] (a) {$(Int \rightarrow Int) \rightarrow Int 
          \rightarrow Int$}
        child {node [left=.5cm, below=1cm, text width=3cm] 
          {$S : \\(t_2 \rightarrow t_1 \rightarrow t_0)  \rightarrow (t_2 \rightarrow t_1) 
            \rightarrow \\ t_2 \rightarrow t_0$}}
        child {node [right=0cm, text width=2cm] {$*: Int \rightarrow Int \rightarrow Int$}}
      }
      child {node [below right=.3cm and 1cm] (b) {$Int \rightarrow Int$}
        child {node {$I: t_0 \rightarrow t_0 \rightarrow t_0$}}
      };
    \end{tikzpicture}
    \label{fig:clbintree}
  \end{subfigure}
  \begin{subfigure}[After]{0.4\linewidth}
    \begin{tikzpicture}[scale=.7, transform shape, 
        level/.style={sibling distance=10mm/#1},
        every edge/.style={draw, thin}
      ]
      \node [circle,draw] (a) {}
      child {node [left=0cm] (b) {$f_1$}}
      child {node [left=.2cm] (c) {\dots} edge from parent[draw=none] }
      child {node [left=.4cm] (e) {$f_n$}}
      child {node [circle,draw, left=1cm] (d) {} edge from parent[very thick]
        child {node [circle, draw, left=.25cm] (l) {} edge from parent[draw=none]
          child {node [left=.5cm] (b) {$f_1$}} 
          child {node [left=.3cm] (c) {\dots} edge from parent[draw=none] }
          child {node [left=.3cm] (e) {$f_n$} edge from parent[very thick]}
          child {node [circle, draw, left=.25cm] (and2) {}
            child {node [circle, draw, left=.25cm](l2) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw] (r2)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
        }
        child {node [circle, draw, right=.25cm] (r) {} edge from parent[draw=none]
          child {node [circle, draw, right=.25cm] (and3) {}
            child {node [circle, draw ](l3) {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
            child {node [circle, draw, right=.2cm ] (r3)  {} edge from parent[draw=none] 
              child {node [above=.5cm] {\vdots} edge from parent[draw=none]}
            }
          }
          child {node [right=.3cm] (b) {$f_1$} edge from parent[very thick]} 
          child {node [right=.4cm] (c) {\dots} edge from parent[draw=none] }
          child {node [right=.6cm] (e) {$f_n$}}
        }
      };

@@

      \path (and3)  edge node[inner sep=0mm,pos=.4] (t1) {} node {} (l3);
      \path (and3)  edge node[inner sep=0mm,pos=.4] (k1) {} node {} (r3);
      \path (t1) edge[bend right] (k1) ;
    \end{tikzpicture}
    \label{fig:andor}
    \end{subfigure}
  \caption{(a) The typed combinator $S * I$ ($f(x)=x^2$) represented as
    binary tree. (b) The space of all typed combinators represented as an
    infinitely deep AND/OR tree. Any specific combinator is a
    sub-binary tree of this AND/OR tree whose leaves consist of only
    the $f_i$ nodes (i.e., the primitive combinators of the grammar).}

@@

\subsection{A stochastic grammar over programs.}
A stochastic grammar over programs represents a distribution over
expressions in $\mathcal{L}$. We represent our grammar as a
multinomial distribution over the primitives in our library. Let $p_1,
\dots, p_m$ be the probabilities associated with primitive combinators
$c_1, \dots, c_m$. Then we define the probability $p(c_i | t) $ of
using $c_i$ at a node whose type is $t$ to be $p_i$ normalized by the
probability mass of all other primitive combinators that could have
been used at that node. Namely, $p(c_i | t) = \frac{p_i}{\sum_{c_j :
    type(c_j) \text{ unifies with } t} p_j}$ if type($c_i)$ unifies
with $t$, 0 otherwise. Note that the distribution thus defined is not
normalizable if we allow expressions of arbitrary depth, but this is
not a practicle concern since we are using the distribution mainly as
a search heuristic.

\subsection{Best-first enumeration of most probable programs.}
In order to explore the frontier of most promising programs, we need a
procedure that enumerates the $N$ most probable programs. There has
been recent interest in this
problem~\cite{DBLP:conf/sfp/Katayama05}\cite{DBLP:conf/aaip/YakushevJ09}. Our
approach is to formulate the problem as a best-first exploration of an
AND/OR tree~\cite{nilsson1982principles}, which we denote
$\mathcal{G}$. A complete program is any binary subtrees of
$\mathcal{G}$ whose leaves coincide with the leaves of
$\mathcal{G}$. The tree structure is shown in
Figure~\ref{fig:andor}. The root is an OR node of the requisite
program type. It's children are the primitives of that type $f_1,
\dots, f_n$, and one AND node with two children. Each child of an AND
node has the same structure as the root node, with modified types: if
the type of an AND node is $t$, then the type of its left child is
$\tau \rightarrow t$ (where $\tau$ is a fresh type variable). Since
the type of the right child is constrained by the subtree rooted at
the left child, we always expand the left child first and use the
target type of this left subtree to mark the type of the right child.

In accordance with the stochastic grammar over programs, we weight all
edges from OR nodes according to the negative log likelihood of
choosing the associated rule from the stochastic grammar, conditioned
on the requesting type. Expanding an AND node has zero cost. In the
case that the stochastic grammar is ill conditioned, we might have an
infinitely deep solution tree whose cost does not increase as we
unfold it; this can occur, for example, if there are no primitives of
the required type since the log probability of expansion to the next
AND node is 0. To avoid this, we simply declare a cost for
expanding the next AND node when it is the only node available.

\subsection{Finding the most compressive set of solutions.}

@@

In our algorithm, we traverse a set of combinator binary trees in
depth first order. We count the occurrences of every subtree. Every
time we encounter a subtree for the \emph{second} time, we designate
it as a primitive in our new grammar. This ensures that every
primitive is used more than once. During our traversal, we do not
descend into trees that have been added already to the new
grammar. Thus, we only count an instance of tree if that instance is
not a subtree of a primitive element of the new grammar.

At the end of this traversal of the solution programs, we have a parse
of the solution set with a new set of primitives and we know how many
times each primitive occurred in the solutions set. To estimate the
probabilities associated with each node production, we again traverse
the solution set, but this time for each node we count which other
primitive elements from the new grammar could have been used. We
estimate the terminal production weight to be the ratio of the number
of times that node was used in the first traversal divided by the
number of times it was used in the second traversal.

@@

%% The choice of an initial grammar for the algorithm will, of course,
%% have a big effect on its performance, and so we would like to
%% characterize this effect. We can divide primitives into roughly two
%% sets; there are low-level primitives, those that operate on concrete
%% types like Bool or Double, and these tend to be domain specific. For
%% example, concrete primitives include $1, 2$, etc., and $f(x) = 3x$. The
%% other class of primitives are abstract primitives, which can be
%% characterized by having arguments whose types are primarily function
%% types, like $(t \rightarrow s), (t \rightarrow s) \rightarrow \text{
%%   Double },$ etc. Of course, we can construct operators that fall in
%% between.

%% The most abstract operators are those that Liang et. al refer to as
%% ``routers''~\cite{liang10programs}. These operators take the place of
%% $\lambda$ abstraction in the general lambda calculus: whereas
%% $\lambda$ abstraction allows us to bind arguments to variables nested
%% deep inside an expression, without them we need to have operators that
%% ``route'' arguments to the appropriate functions in a nested
%% expression.

%% \begin{definition}[Router~\cite{liang10programs}] A \emph{router} is a combinator
%% represented by a finite sequence of elements from $\{B, C, S\}$. The
%% behavior of router $r$ is given by
%% \[
%% (r \, f \, g \, x_1 \dots x_{|r|}) = (f \, x_{i_1} \dots x_{i_n}) (g
%% \, x_{j_1} \dots x_{j_n}),
%% \]
%% where $i_1 < \dots < i_n$ are indices $i$ such that $r_i \in \{C, S\}$
%% and where $j_1 < \dots < j_n$ are indices $j$ such that $r_j \in \{B,
%% S\}$. Let $\mathcal{R}_k$ be the set of all routers of length equal to
%% $k$. Let $\mathcal{R}_{\leq k}$ be the set of all routers of length
%% less than or equal to $k$.
%% \end{definition}

%% To analyze the behavior of our algorithm, we will vary the parameters
%% of the problem along two axes while keeping the learning tasks the
%% same. Along the first axis, we vary the initial library to which our
%% learner has access with respect to the its abstractness (i.e., number
%% of routers) and its concreteness (i.e. number of concrete
%% primitives). Along the second axis we vary the size of the search
%% frontier.

Each task is to predict the value of an unknown polynomial on the
inputs $0, \dots, 10$. The set of tasks corresponds to the set of
polynomials $\{ax^2 + bx + c | a, b, c \in \{0, \dots, 10 \}\}$. The
choice of primitives is going to make a big difference in performance;
we chose to include the basic combinators I, S, B, C and four
arithmetic primitives, $1$, $0$, $*$, and $+$. The choice of initial
weights makes a large difference as well, as it determines the
relative ordering of combinators in the enumeration step. To get a
general sense of the algorithm's performance, we set the intial
weights to be slightly different on each run, jittering them around a
uniform weighting.

In Figure~\ref{fig:symreg}, we show performance results as we vary the
frontier size. Each individual plot shows how changing the frontier
size changes the number of tasks the algorithm identifies correctly
over the course of fifteen algorithm iterations. %% In addition, we have a
%% baseline for each curve, visits the same number of expressions as the
%% corresponding algorithm run (i.e. frontier size $\times$ $\#$ of
%% iterations) but all at once; it does not do any learning. Thus, this
%% baseline serves as a brute-force control for the learning aspect of
%% the algorithm.

\begin{figure}
\begin{subfigure}[Before]{0.45\linewidth}
\includegraphics[width=\linewidth]{figures/learningCurves2.pdf}
\end{subfigure}
\begin{subfigure}[After]{0.45\linewidth}
\includegraphics[width=\linewidth]{figures/performanceVsFrontierSize2.pdf}
\end{subfigure}
\caption{(a) Average learning curves as we vary the frontier size. (b)
  Percent of tasks solved at iteration 15 as a function of frontier
  size.}
\label{fig:symreg} 
\end{figure}
